---
title: "Mini Data Analysis Milestone 2"
output: 
  github_document:
    fig_width: 10
    fig_height: 10
    dev: png
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by:

-   Making summary tables and graphs
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.*

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission.

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Task 1: Process and summarize your data

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (1 point)

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->

1.  *How are tree species distributed across various neighborhoods and streets in Vancouver?*

2.  *Which tree species is predominant in each neighborhoods and streets in Vancouver*

3.  *Were there particular periods during which tree planting significantly increased in Vancouver*

4.  *Is there a discernible correlation between the diameter or height of trees and factors like the presence of a root barrier*

### 1.1.1 (Some comments)

1.  How many different tree species are there in each neighborhood and street?

2.  The second research question has been modified for enhanced graphical representation and clarity.

3.  Which tree species were primarily planted during these annual periods?

4.  The fourth research question has been refined for clearer graphical representation: I will initially examine any correlations between the diameter or height of trees and the presence of a root barrier. Subsequently, I will explore other factors.

<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks:

8.  Create a graph that has at least two geom layers.
9.  Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->

```{r}
# To have a general understanding of  vancouver_trees data set, 
# I will firstly use glimpse() to check its overall structure
glimpse(vancouver_trees)
vancouver_trees_age <- vancouver_trees %>%
  mutate(year_plant = as.numeric(format(date_planted, "%Y")), tree_age = 2023 - year_plant)
vancouver_trees_age
```

## **Research Question 01:** How are tree species distributed across various neighborhoods and streets in Vancouver?

**Summarizing task: Task 4: Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`**.

-   I will analyze the number and percentages of trees by species in each neighborhood and street.
-   Additionally, I have calculated the counts and proportions of each species within every neighborhood and street.
-   This analysis will be instrumental in addressing my research question, providing a clear perspective on the distribution of tree species across different neighborhoods and streets.

**Graphing task: Task 8: Create a graph that has at least two geom layers**.

-   I will utilize geom_bar to depict the counts and proportions of species within each neighborhood.
-   Additionally, I have visualized the distribution of tree species by street.
-   This is instrumental in addressing my research question, as it provides a clear representation of how tree species are distributed across different neighborhoods and streets.

```{r}
#####################Summarizing task: Task 4  ############################
# summarizing how many species there are in this data set
# total number of tree species in  Vancouver
total_species_count <- n_distinct(vancouver_trees$species_name)
# there are 238 tree species in the City of Vancouver
total_species_count

# Group tree for each neighborhood of each tree species
tree_species_per_neighborhoods <- vancouver_trees %>% group_by(neighbourhood_name, species_name)
# Check the number of trees for each tree species in every neighborhood
tree_species_per_neighborhoods_prop <-tree_species_per_neighborhoods %>% summarise(total_trees = n()) %>% mutate(prop = total_trees/sum(total_trees))
tree_species_per_neighborhoods_prop
# Check the number of species in every neighborhood
total_tree_species_per_neighborhoods <-  vancouver_trees %>% group_by(neighbourhood_name) %>% 
    summarize(total_species = n_distinct(species_name)) %>% 
    mutate(prop = total_species/total_species_count)
total_tree_species_per_neighborhoods

# Group tree for each street of each tree species
tree_species_per_streets <- vancouver_trees %>% group_by(std_street, species_name)
# Check the number of tree for each tree species in every street
tree_species_per_street_prop <-tree_species_per_streets %>% summarise(total_trees = n()) %>% mutate(prop = total_trees/sum(total_trees))
tree_species_per_street_prop
# Check the number of species in every street
total_tree_species_per_streets <-  vancouver_trees %>% group_by(std_street) %>% 
    summarize(total_species = n_distinct(species_name)) %>% 
    mutate(prop = total_species/total_species_count)
total_tree_species_per_streets

##################### Graphing task: Task 8 ############################
ggplot(total_tree_species_per_neighborhoods, aes(x = neighbourhood_name, y = total_species)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) + 
  geom_text(aes(label=total_species), vjust=-0.05) + # Adds a text layer to display the number of species
  labs(title = "Number of Tree Species per Neighborhood",
       y = "Total Number of Species",
       x = "Neighborhood Name") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Distribution of Tree Species per Street
ggplot(total_tree_species_per_streets, aes(x = total_species)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "skyblue", alpha = 0.7) + 
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Tree Species per Street",
       x = "Number of Tree Species",
       y = "Density") +
  theme_minimal()

```



## **Research Question 02:** Which tree species is predominant in each neighborhoods and streets in Vancouver?

**Summarizing task: Task 2: Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`**.

-  I will determine the number of predominant tree species in each neighborhood and street.
- This information is crucial for my research, as it provides a clear insight into which tree species are most prevalent in various neighborhoods and streets of Vancouver.

**Graphing task: Task 7: Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best**.

-  I will employ geom_bar to depict the dominant species in each neighborhood.
-  This visualization is crucial for addressing my research query as it directly showcases the dominant species in every neighborhood.
-  A larger bin size is preferable as it condenses the graph, offering a clearer view to users.

```{r}
#####################Summarizing task: Task 3  ############################
predominant_species_of_each_neighbourhood <- vancouver_trees %>%
  group_by(neighbourhood_name, species_name) %>%
  summarise(count = n()) %>%
  arrange(neighbourhood_name, -count) %>%
  slice(1) %>%
  select(neighbourhood_name, species_name, count)
predominant_species_of_each_neighbourhood

predominant_species_of_each_street <- vancouver_trees %>%
  group_by(std_street, species_name) %>%
  summarise(count = n()) %>%
  arrange(std_street, -count) %>%
  slice(1) %>%
  select(std_street, species_name, count)
predominant_species_of_each_street

##################### Graphing task: Task 8 ############################
# Histogram 1: Small bin width
hist1 <- ggplot(predominant_species_of_each_neighbourhood, aes(x = neighbourhood_name, y = count, label = species_name)) +
  geom_bar(stat = 'identity', fill = "skyblue", alpha = 0.7, width = 0.25) +
  geom_text(aes(label=species_name), vjust=-0.5, size = 2) +
  labs(title = "Histogram 1: Small Bin Width",
       x = "Neighbourhood Name",
       y = "Count of Predominant Species Trees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Histogram 2: Medium bin width
hist2 <- ggplot(predominant_species_of_each_neighbourhood, aes(x = neighbourhood_name, y = count, label = species_name)) +
  geom_bar(stat = 'identity', fill = "lightgreen", alpha = 0.7, width = 0.5) +
  geom_text(aes(label=species_name), vjust=-0.5, size = 2) +
  labs(title = "Histogram 2: Medium Bin Width",
       x = "Neighbourhood Name",
       y = "Count of Predominant Species Trees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Histogram 3: Large bin width
hist3 <- ggplot(predominant_species_of_each_neighbourhood, aes(x = neighbourhood_name, y = count, label = species_name)) +
  geom_bar(stat = 'identity', fill = "lightpink", alpha = 0.7, width = 0.7) +
  geom_text(aes(label=species_name), vjust=-0.5, size =2) +
  labs(title = "Histogram 3: Large Bin Width",
       x = "Neighbourhood Name",
       y = "Count of Predominant Species Trees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

## Create a slice of predominant_species_of_each_street of 20 rows
predominant_species_of_each_street_first_20 <- head(predominant_species_of_each_street, 20)

# Histogram 4: Small bin width
hist4 <-ggplot(predominant_species_of_each_street_first_20, aes(x = std_street, y = count, label = species_name)) +
  geom_bar(stat = 'identity', fill = "skyblue", alpha = 0.7, width = 0.25) +
  geom_text(aes(label=species_name), vjust=-0.5, size = 2) +
  labs(title = "Histogram 1: Small Bin Width",
       x = "Street Name",
       y = "Count of Predominant Species Trees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Histogram 5: Medium bin width
hist5 <- ggplot(predominant_species_of_each_street_first_20, aes(x = std_street, y = count, label = species_name)) +
  geom_bar(stat = 'identity', fill = "lightgreen", alpha = 0.7, width = 0.5) +
  geom_text(aes(label=species_name), vjust=-0.5, size = 2) +
  labs(title = "Histogram 2: Medium Bin Width",
       x = "Street Name",
       y = "Count of Predominant Species Trees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Histogram 6: Large bin width
hist6 <- ggplot(predominant_species_of_each_street_first_20, aes(x = std_street, y = count, label = species_name)) +
  geom_bar(stat = 'identity', fill = "lightpink", alpha = 0.7, width = 0.7) +
  geom_text(aes(label=species_name), vjust=-0.5, size = 2) +
  labs(title = "Histogram 3: Large Bin Width",
       x = "Street Name",
       y = "Count of Predominant Species Trees") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

print(hist1)
print(hist2)
print(hist3)
print(hist4)
print(hist5)
print(hist6)
```



## **Research Question 03:** Were there particular periods during which tree planting significantly increased in Vancouver?

**Summarizing task: Task 1: Compute the range, mean, and two other summary statistics of one numerical variable across the groups of one categorical variable from your data**. 

-  I will determine the annual tree planting count in Vancouver.
-  Subsequently, I will compute the range, mean, standard deviation, and median of trees planted, categorized by street_side_name.
-  This analysis will be instrumental in addressing my research question, providing a clear perspective on Vancouver's tree planting efforts and revealing the contribution of each street_side annually.

**Graphing task: Task 8:Create a graph that has at least two geom layers**.

-  I will employ geom_point to depict the annual tree planting activities in Vancouver.
-  This visualization will be instrumental in addressing my research question, offering a clear representation of yearly tree planting in Vancouver.

```{r}
#####################Summarizing task: Task 3  ############################
# Remove rows where Year_Column is NA
vancouver_trees_age_clean <- vancouver_trees_age %>% filter(!is.na(year_plant))

# Group by Year_Column to get the count of trees planted each year
tree_planting_per_year <- vancouver_trees_age_clean %>%
  group_by(year_plant) %>%
  summarise(num_trees_planted = n())
tree_planting_per_year

tree_planting_with_street_side_per_year <- vancouver_trees_age_clean %>%
  group_by(year_plant, street_side_name) %>%
  summarise(num_trees_planted = n())
tree_planting_with_street_side_per_year

summary_stats <- tree_planting_with_street_side_per_year %>%
  summarise(
    range = max(num_trees_planted) - min(num_trees_planted),
    mean = mean(num_trees_planted),
    median = median(num_trees_planted),
    sd = sd(num_trees_planted)
  )
summary_stats

# Plot
p1 <- ggplot(tree_planting_per_year, aes(x = year_plant, y = num_trees_planted)) +
  geom_line(color = "blue", size = 1) + 
  geom_point(color = "red", size = 2) +
  geom_text(aes(label = year_plant), vjust = -0.5, size = 3) +
  labs(title = "Tree Planting Trend in Vancouver Over the Years",
       x = "Year of Planting",
       y = "Number of Trees Planted") +
  theme_minimal()

print(p1)
```



## **Research Question 04:** Does the tree with a root barrier have a long diameter and higher height in general?

**Summarizing task: Task 1: Compute the range, mean, and two other summary statistics of one numerical variable across the groups of one categorical variable from your data**.

-  I need to analyze the average diameter and height of trees with a root barrier in comparison to those without one using the vancouver_trees data set.
- This is advantageous as it provides a statistical perspective on the height and diameter associated with the root barrier.

**Graphing task: Task 8: Create a graph that has at least two geom layers**. 

- I will create a bar chart that displays the average diameter and height for each root barrier group.
- By visualizing these metrics, users can make informed decisions about implementing or re-evaluating root barriers, considering their potential impact on tree growth.

```{r}
#####################Summarizing task: Task 3  ############################
# Remove rows where Year_Column is NA
# Summarize data by presence of root barrier
tree_stats <- vancouver_trees %>%
  group_by(root_barrier) %>%
  summarise(
    mean_diameter = mean(diameter, na.rm = TRUE),
    mean_height = mean(height_range_id, na.rm = TRUE)
  )
tree_stats

# Melt the data for easier plotting
tree_stats_long <- pivot_longer(tree_stats, c(mean_diameter, mean_height), names_to = "measure", values_to = "value")
tree_stats_long
# Plot
ggplot(tree_stats_long, aes(x = root_barrier, y = value, fill = measure)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.2f", value)), position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Comparison of Mean Diameter and Height with/without Root Barrier",
       y = "Value",
       x = "Root Barrier Presence") +
  theme_minimal()
```

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

### Research Question 1: How are tree species distributed across various neighborhoods and streets in Vancouver?

-   I've successfully mapped out the distribution of tree species across neighborhoods and streets in Vancouver by creating a detailed tibble.
-   Visual representations, such as plots, further illustrate this distribution, providing clarity on concentrations in specific neighborhoods and streets.
-   To delve deeper, a geographical heat-map on a map of Vancouver could provide a visual representation of the distribution of tree species by location.

### Research Question 2: Which tree species is predominant in each neighborhoods and streets in Vancouver?

-   By analyzing the data, I have identified the most predominant tree species in every neighborhood and street in Vancouver.
-   The visual plots created offer a snapshot of the most frequently found tree species across these areas.
-   Moving forward, it would be insightful to identify which neighborhoods lack diversity in tree species, and which species are exclusive to specific areas.

### Research Question 3: Were there particular periods during which tree planting significantly increased in Vancouver?

-   The data has been organized year-wise to highlight the number of trees planted annually in Vancouver.
-   The plots vividly depict fluctuations in tree planting activities over the years, pointing out significant surges.
-   A deeper analysis could involve understanding the reasons behind these surges. This could be tied to city policies, environmental campaigns, or other socio-economic factors.

### Research Question 4: Is there a discernible correlation between the diameter or height of trees and factors like the presence of a root barrier, street side, curb or ages(date of plant)

-   Preliminary data suggests a correlation between tree dimensions (diameter and height) and the presence of root barriers.
-   The visuals generated support this observation by presenting noticeable differences in dimensions in trees with and without root barriers.
-   As a subsequent step, I aim to explore potential correlations between tree dimensions and other factors such as street side, proximity to the curb, and the age of the tree. Regression models or correlation matrices might be used to further investigate these relationships.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

```{r}
#select 8 variables
glimpse(vancouver_trees)
tidy_trees <- vancouver_trees %>% 
  select(c(tree_id, civic_number, std_street, cultivar_name, common_name, date_planted, longitude, latitude))
glimpse(tidy_trees)
```

Based on the definition above, I will pick **tree_id**, **civic_number**, **std_street**, **cultivar_names**, **common_name**, **date_planted**, **longitude**, **latitude** 8 variables.

Overall, the data adheres to the principles of tidy data. However, there are some missing values (NAs) within the std_street, cultivar_names, date_planted, longitude, latitude categories. Nonetheless, the data set consistently ensures that each row represents an observation, each column signifies a variable, and each cell contains a single value

All of the columns are variables, and all of their contents are values. Hence, tidy data. <!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

```{r}
# Convert the tidy data into untidy data by combining the species_name and cultivar_name
tidy_to_untidy <- vancouver_trees %>% pivot_longer(cols = c("species_name", "cultivar_name"), 
               names_to  = "types_of_taxonomy", 
               values_to = "name")
print(tidy_to_untidy)

# Convert the untidy data into tidy data
untidy_to_tidy_original <-tidy_to_untidy %>% 
  pivot_wider(names_from  = "types_of_taxonomy", values_from = "name")
print(untidy_to_tidy_original)

#Part B: Untidy Data to Tidy Data 
tidy_to_drop_na_tidy_data <- vancouver_trees %>% drop_na()
print(tidy_to_drop_na_tidy_data)
```

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  *Were there particular periods during which tree planting significantly increased in Vancouver*
2.  *Is there a discernible correlation between the diameter or height of trees and factors like the presence of a root barrier*

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->

1.  Investigating Tree Planting Trends in Neighborhoods The initial investigation revealed intriguing patterns in tree planting, with certain years witnessing a more pronounced inclination towards afforestation. This piqued my curiosity: while the broader trend was evident, did individual neighborhoods mirror this pattern or did they exhibit unique trends of their own? Understanding the micro-trends at the neighborhood level can offer insights into localized factors or initiatives that might be driving or inhibiting tree planting. This could be instrumental in tailoring community-specific strategies to promote green initiatives.

2.  Factors Influencing Tree Growth While the impact of root barriers on tree diameter and height has been established, it's essential to delve deeper and uncover other potential factors influencing tree growth. Trees don't just grow in isolation; they are influenced by a myriad of environmental, biological, and human-induced factors. By identifying these factors, we can pinpoint neighborhoods that offer conducive environments for tree growth. This knowledge can guide urban planning decisions, ensuring that trees planted today grow tall and robust, contributing to a greener urban landscape in the future. <!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.)

<!--------------------------- Start your work below --------------------------->

### Further Investigation on Research Question 01

```{r}
# Group by Year_Column and neighborhood to get the count of trees planted each year for each neighborhood
tree_planting_per_year_per_neighbourhood <- vancouver_trees_age_clean %>%
  group_by(year_plant, neighbourhood_name) %>%
  summarise(tree_count = n())
tree_planting_per_year_per_neighbourhood

summary_stats <- tree_planting_per_year_per_neighbourhood %>%
  group_by(neighbourhood_name) %>%
  summarise(
    tree_count_max = max(tree_count),
    tree_count_min = min(tree_count),
    tree_count_range = tree_count_max - tree_count_min,
    max_tree_year_plant = year_plant[which.max(tree_count)],
    min_tree__year_plant = year_plant[which.min(tree_count)],
    tree_count_mean = mean(tree_count),
    tree_count_std_dev = sd(tree_count),
    tree_count_median = median(tree_count)
  )

summary_stats

p1 <- ggplot(tree_planting_per_year_per_neighbourhood, aes(x = year_plant, y = tree_count, color = neighbourhood_name)) +
  geom_line() +
  labs(title = "Tree Planting Trends by Neighborhood in Vancouver",
       x = "Year",
       y = "Number of Trees Planted") +
  theme_minimal() +
  theme(legend.position="none")  # you can remove this line if you want the legend to be displayed
print(p1)

```


### Further Investigation on Research Question 02

```{r}
# Is there a discernible correlation between the diameter or height of trees and factors like the presence of street side, curb or ages(date of plant)
tree_growth_stats <- vancouver_trees_age %>%
  select(tree_id, curb, street_side_name, tree_age, diameter,height_range_id)

# stat for curb
tree_growth_curb_stats <- tree_growth_stats %>% group_by(curb) %>%
  summarise(
    mean_diameter = mean(diameter, na.rm = TRUE),
    mean_height = mean(height_range_id, na.rm = TRUE)
  )
tree_growth_curb_stats

# stat for street_side_name
tree_growth_street_side_stats <- tree_growth_stats %>% group_by(street_side_name) %>%
  summarise(
    mean_diameter = mean(diameter, na.rm = TRUE),
    mean_height = mean(height_range_id, na.rm = TRUE)
  )
tree_growth_street_side_stats

# stat for year_plant
tree_growth_age_stats <- tree_growth_stats %>% group_by(tree_age) %>%
  summarise(
    mean_diameter = mean(diameter, na.rm = TRUE),
    mean_height = mean(height_range_id, na.rm = TRUE)
  )
tree_growth_age_stats

# Ploting
# Melt the data for easier plotting to curb
tree_stats_growth_curb_long <- pivot_longer(tree_growth_curb_stats, c(mean_diameter, mean_height), 
                                names_to = "measure", values_to = "value")
tree_stats_growth_curb_long
# Plot
ggplot(tree_stats_growth_curb_long, aes(x = curb, y = value, fill = measure)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.2f", value)), position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Comparison of Mean Diameter and Height with/without Root Barrier",
       y = "Value",
       x = "Curb Presense") +
  theme_minimal()

# Melt the data for easier plotting to street name
tree_stats_growth_street_side_long <- pivot_longer(tree_growth_street_side_stats, c(mean_diameter, mean_height), 
                                names_to = "measure", values_to = "value")
tree_stats_growth_street_side_long
# Plot
ggplot(tree_stats_growth_street_side_long, aes(x = street_side_name, y = value, fill = measure)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.2f", value)), position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Comparison of Mean Diameter and Height with/without Root Barrier",
       y = "Value",
       x = "Street Side") +
  theme_minimal()

# Melt the data for easier plotting to street name
tree_stats_growth_ages_long <- pivot_longer(tree_growth_age_stats, c(mean_diameter, mean_height), 
                                names_to = "measure", values_to = "value")
tree_stats_growth_ages_long
# Plot
ggplot(tree_stats_growth_ages_long, aes(x = tree_age, y = value, fill = measure)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.2f", value)), position = position_dodge(width = 0.9), vjust = -0.5, size = 1.75) +
  labs(title = "Comparison of Mean Diameter and Height with/without Root Barrier",
       y = "Value",
       x = "Tree ages until 2023") +
  theme_minimal()
```

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: *Is there a discernible correlation between the diameter or height of trees and factors like the presence of a root barrier(curb, ages, street side name)*

**Variable of interest**: *tree_growth_stats*

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->

```{r}
# Fitting lm model on diameter based on tree_age
age_plant_diameter_lm <- lm(formula = diameter ~ tree_age , data = tree_growth_stats)
summary(age_plant_diameter_lm)

# This is just for exploration on diameter based on height_range_id (not related too much to this specific question)
# But this gives a idea how height_range_id would relates to diameter
height_range_diameter_lm <- lm(formula = diameter ~ height_range_id , data = tree_growth_stats)
summary(height_range_diameter_lm) 
```

<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->

**I will produce a p-value to determine if age is a eligible predictor of the tree diameter.** 
**I will produce a p-value to determine if height range id is a eligible predictor of the tree diameter.**

```{r}
lm_tidy_age_plant_diameter <- broom::tidy(age_plant_diameter_lm)
age_plant_diameterp_values <- lm_tidy_age_plant_diameter$p.value
summary(age_plant_diameterp_values) #the p-value is greater than zero but very small

lm_tidy_height_range_id_diameter <- broom::tidy(height_range_diameter_lm)
height_range_diameter_p_values <- lm_tidy_height_range_id_diameter$p.value
summary(height_range_diameter_p_values) #the p-value is greater than zero but very small
```

<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->

**The varialbe is total_tree_species_per_neighborhoods in Task 1 Research Question 01**

```{r}
# install.packages("here") 
library(here)
library(readr)
# Define the directory path
output_path <- here("output")
# Create the directory if it doesn't exist
if (!dir.exists(output_path)) {
  dir.create(output_path)
}
csv_save_path <- here("output/total_tree_species_per_neighborhoods.csv")
write_csv(total_tree_species_per_neighborhoods, file = csv_save_path)
```

<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->

```{r}
# Define the model path
age_model_output_path <- here("output", "age_plant_diameter_lm.rds" )
height_model_output_path <- here("output", "height_range_diameter_lm.rds" )

# write as RDS
saveRDS(age_plant_diameter_lm, file = age_model_output_path)
saveRDS(height_range_diameter_lm, file = height_model_output_path)

# read RDS
age_plant_diameter_lm_rds <- readRDS(age_model_output_path)
age_plant_diameter_lm_rds
height_range_diameter_lm_rds <- readRDS(height_model_output_path)
height_range_diameter_lm_rds

# Use identical to check if the read version rds is same as the original one in R format
identical(age_plant_diameter_lm_rds, age_plant_diameter_lm)
identical(height_range_diameter_lm_rds, height_range_diameter_lm)
```

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors.

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone.

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files.
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2.

### Attribution

Thanks to Victor Yuan for mostly putting this together.
